% arara: pdflatex: { shell: yes, interaction: nonstopmode }
% arara: pythontex: {verbose: yes, rerun: modified }
% arara: pdflatex: { shell: yes, interaction: nonstopmode }
% arara: clean: { extensions: [ aux, blg, idx, ilg, ind, log, out, pytxcode, rel, toc ] }
% !arara: clean: { files: [ ans.tex, hint.tex] }

\documentclass[queueing-book.tex]{subfiles}
%\externaldocument{queueing-book}
\input{header}

\begin{document}
\section{Little's Law}
\label{sec:littles-law}

There exists a fascinating connection between the average sojourn time of a job in a (queueing) system and the long-run time-average number of jobs within that system.
This relationship is known as \emph{Little's law} and stands as one of the most useful results in any input-output system, including queueing systems.
To understand the essence of Little's law, let's consider an illustrative example: imagine a highway segment between points A and B spanning 100 km.
Every minute, one car enters the highway, and each car maintains a constant speed of 50 km/h.
Now, let's determine the number of cars on that highway segment.
Since each car spends 2 hours on the highway,  there must be 120 cars traversing the distance at any moment in time.
Stated as a formula, if items arrive at rate $\lambda$ and each item spends a time $\E J$ time in the system, then the total number of items in the system is $\E L = \lambda \E J$.

Our objective in this section is to demonstrate that this law also applies to stochastic systems, given a few simple conditions.


\newthought{Let's begin by} expressing the sojourn time of a particular job, denoted as~$k$, in terms of an indicator function\sidenote{The indicator $\1{A} = 0$ if the event $A$ occurs, and $0$ otherwise.}.
When job~$k$ is present in the system at time~$s$, then $\1{A_{k}\leq s < D_{k}}$ is 1, and 0 otherwise.
Consequently, the integral $\int_{0}^{t} \1{A_{k}\leq s < D_{k}} \d s$ represents the total time that job~$k$ spends in the system until time~$t$.
\sidenote{A helpful analogy is to imagine that a job pays~$1$ Euro for each unit of time spent in the system, making this integral equivalent to the total payment made by job~$k$ up to time~$t$.}
Hence, we can express the sojourn time of the~$k$th job as follows:
\begin{equation*}
\J_k = \int_0^\infty \1{A_k \leq s < D_k},\d s.
\end{equation*}


Furthermore, it is evident that $\L(s) = \sum_{k=1}^\infty \1{A_k \leq s < D_k}$ represents the number of jobs in the system at time~$s$.
By combining these two facts, the LHS of the equation
\begin{equation}\label{eq:little:2}
\int_{0}^{t}\L(s) \d s = \int_{0}^{t }\sum_{k=1}^\infty \1{A_k \leq s < D_k} \d s = \sum_{k=1}^\infty \int_{0}^{t }\1{A_k \leq s < D_k} \d s
\end{equation}
can be interpreted as the total time that jobs have spent in the system up to time~$t$.\sidenote{Alternatively, it's the total payment made by the jobs up to time~$t$.}


\newthought{Consider a specific} time~$T$ at which the system is empty. At this point, any job $k\leq A(T)$ must have left the system, implying that $ \J_k = \int_0^T \1{A_k \leq s < D_k}\,\d s$ when $k\leq A(T)$, therefore,
\begin{equation}\label{eq:111}
 \int_0^T \L(s)\, \d s  = \int_0^T \sum_{k=1}^{A(T)} \1{A_k \leq s < D_k}\, \d s
 = \sum_{k=1}^{A(T)}\int_0^T \1{A_k \leq s < D_k}\, \d s = \sum_{k=1}^{A(T)} \J_k.
\end{equation}

Suppose now there exists an infinite sequence of  times $\{T_i\}$ at which the system is empty, i.e., $A(T_i) = D(T_i)$, and such that $T_i\to\infty$. In that case we can deduce that
\begin{equation*}
 \frac 1 {T_i} \int_0^{T_i} \L(s)\, \d s = \frac{A(T_i)} {T_i} \frac{1}{A(T_i)} \sum_{k=1}^{A(T_i)} \J_k.
\end{equation*}
By taking the limit $i\to\infty$ on both sides, provided the limits exist, we arrive at \recall{Little's law}:\sidenote{Use the physical dimensions of the components of Little's law to check that $\E{\J} \neq \lambda \E{\L}$.}
\begin{equation*}
 \E \L = \lambda \E \J.
\end{equation*}
It is important to note that  Little's law does  not necessarily hold at an arbitrary moment in time, instead, it expresses a relation between \emph{averages}.


\newthought{In the above} proof of Little's law we assumed that there is an increasing sequence of epochs $\{T_k, k=0,1,\ldots\}$ at which the system is empty.
However, in many practical queueing situations the system is never empty.
It can be proven that Little's law still holds under the weaker condition of rate-stability.




\begin{exercise}\label{ex:37}
Think of all servers of the (stable) $G/G/c$ queue as being put in a box.
Jobs enter the box at rate $\lambda$ and stay $\E S$ in this box.
Use Little's law to conclude that the time-average number of busy server is $\E{\Ls} = \lambda \E S$.
\begin{hint}
Recall that $\lim_{t\to\infty} t^{-1}\int_0^t \Ls(s)\d s$ is the time-average number of servers busy up to~$t$.
\end{hint}
\begin{solution}
Since the queue is stable by assumption, jobs depart at rate $\lambda$, hence any job enters and leaves the box that contains the~$c$ servers.
The average time a job spends in the box is $\E S$.
By Little's law, the average number in the box is $\E\Ls = \lambda \E S$.
\end{solution}
\end{exercise}

\begin{exercise}\label{ex:44}
Consider the $M/M/1$ queue when $\lambda=5/h$ and $\mu=6/h$.
Suppose that, at the moment you join the system, the number of customers in the system is 10.
What is your expected waiting time?
\begin{solution}
When you arrive, you first have to wait for the job in service to finish and then for the 9 jobs in queue.
By the memoryless property, at the moment you arrive, the remaining service time of the job in service is still $\E S$ Thus, you have to wait $10/\mu = 5/3 h \neq 5/6 h$.
\end{solution}
\end{exercise}


% \begin{exercise}\label{ex:59}
% Derive \cref{eq:111}.
% \begin{hint}
%  Substitute the definition of $\L(s) = \sum_{k=1}^{A(T)}\1{A_k \leq s < D_k}$ for $s\leq T$ in  the LHS, then reverse the integral and summation.
% \end{hint}
% \begin{solution}
% \begin{equation*}
%  \begin{split}
%  \int_0^T \L(s)\, \d s & = \int_0^T \sum_{k=1}^{A(T)} \1{A_k \leq s < D_k}\, \d s \\
% & = \sum_{k=1}^{A(T)}\int_0^T \1{A_k \leq s < D_k}\, \d s = \sum_{k=1}^{A(T)} \J_k.
%  \end{split}
% \end{equation*}
% \end{solution}
% \end{exercise}

\begin{exercise}
Let's take `life' itself as an input-output system, and ask how many people are born and die in the Netherlands per week.
For ease, assume that in 2023 the Netherlands has 16 M inhabitants, and people have an expected lifetime (i.e., the time they remain in the system called `life') of 80 years.
With Little's law we conclude that the rate $\lambda$ into the system (i.e., people born) is $16 M/80 = 0.2 M $ per year.
Again for ease, if a year has 50 weeks, there are 4000 births per week.
If the size of the population remains about constant, also around 4000 people die per week.

As it turns out, the death rate in 2023 is about 2900 per week.
This is much lower than the above estimate, and a more precise computation does not close the gap.
Can you explain this paradoxical difference?

\begin{hint}
What happens if the population increases?
\end{hint}
\begin{solution}
Let's make a very simple model to clarify the problem.
We start with an initial population size of~$A$.
Then assume that each and every individual dies after exactly 80 years (and does not live a day longer).
Moreover, assume that the population increases by $1\%$ per year.
Then, after 79 years, the size of the population is $A \sum_{k=0}^{79}(1.01)^{k}$.
If we assume (as we do when we apply Little's law) that each year contains the same amount of people, then the number of people per year equals:
\begin{equation*}
\frac{A}{80}\sum_{k=0}^{79} (1.01)^k = \frac{A}{80} \frac{(1.01)^{80}}{1.01-1} = 1.5 A.
\end{equation*}
However, after 80 years, just~$A$ people die, not $1.5 A$.
Interestingly, $4000/1.5 = 2700$, which is quite a bit closer to 2900 per week.

In conclusion, we should apply Little's law with caution; in particular, when we just consider stretches of time in which we know that the population increases (or decreases) on average, Little's law does not hold.

\end{solution}
\end{exercise}

\begin{exercise}\label{ex:l-220}
Explain that $\sum_{k=1}^{A(t)} \J_k \geq \int_0^t \L(s) \d s \geq \sum_{k=1}^{D(t)} \J_k$.
\begin{solution}
 Intuitively, the left term is all the work that arrived up to time~$t$, the middle term is all the work that has been processed, and the right term all the work that left.
 Any job that is half-way its service counts for full at the left, for half in the middle expression, and not in the right.

 More formally, for any job~$k$ and time~$t$, we have $\J_k \1{A_k \leq t} \geq \int_0^t \1{A_k \leq s < D_k } \d s \geq \J_k \1{D_k \leq t}$. (To see this, fix~$k$, and check the three cases $t < A_k, A_k \leq t < D_k, D_k < t$.) Then,
 \begin{equation*}
 \sum_{k=1}^\infty \J_k \1{A_k \leq t} \geq \int_0^t \sum_{k=1}^\infty \1{A_k \leq t < D_k} \d s \geq \sum_{k=1}^\infty \J_k \1{D_k \leq t}.
 \end{equation*}
 Finally, note that $ \sum_{k=1}^\infty \J_k \1{A_k \leq t} = \sum_{k=1}^{A(t)} \J_k$ and $ \sum_{k=1}^\infty \J_k \1{D_k \leq t} = \sum_{k=1}^{D(t)} \J_k$, and use the definition of $\L(s)$.
\end{solution}
\end{exercise}

\begin{exercise}[Continuation of \cref{ex:l-220}]\label{ex:l-221}
Take suitable limits to show that
$\lambda \E\J \geq \E \L \geq \delta \E\J$.
 Where do you need the strong law of large numbers?
\begin{solution}
 \begin{equation*}
 \lim_{t\to\infty} \frac{A(t)}{t}\frac 1{A(t)}\sum_{k=1}^{A(t)} \J_k \geq \lim_{t\to \infty} \frac 1 t \int_0^t \L(s) \d s \geq \lim_{t\to\infty} \frac{D(t)}{t}\frac 1{D(t)} \sum_{k=1}^{D(t)} \J_k.
 \end{equation*}
 We use the strong law of large numbers to conclude that the limits converges to $n^{-1} \sum_{k=1}^n \J_k \to \E \J$, and we assume that $\{\J_k, k\geq N\}$ is a sequence of iid random variables for~$N$ sufficiently large.
\end{solution}
\end{exercise}


\input{trailer}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
