\documentclass[queueing-book.tex]{subfiles}
%\externaldocument{queueing-book}
\input{header}

\begin{document}


\section{\texorpdfstring{$M/G/1$}{MG1} Queue: Expected Waiting Time}
\label{sec:mg1}


When the service times are not really well approximated by the exponential distribution, the $M/G/1$ queue may be a more suitable model than the $M/M/1$ queue.
In this section we derive with sample path arguments the fundamentally important Pollaczek-Khinchine (PK) formula for the average waiting time of an $M/G/1$ queue.
%At the end of the section we relate Sakasegawa's formula to the PK formula.


\newthought{We start with} the simple observation that before an arriving job gets access to the server, it first has to wait until the job in service (if any) completes, and then it has wait for the queue to be cleared.
From PASTA it follows that $\E{\W} = \E{S_r} + \E{\Q} \E S$, where $\E{S_r}$ is the (time-)average remaining service time of the job in service.
With Little's law $\E \Q = \lambda \E\W$ and writing $\rho = \lambda \E S$, we find that\sidenote{Compare the derivation for $\E W$ for the $M^{X}/M/1$ queue, it's similar.}
\begin{equation*}
 \E{\W} = \frac{\E{S_r}}{1-\rho}.
\end{equation*}

To make further progress we need to find an expression for $\E{S_r}$ when~$S$ is some generally distributed rv.\sidenote{For the $M/M/1$ queue, $\E{S_r} = \rho\E S$, but not fora the $M/G/1$ queue.}
For this, we use the renewal reward theorem, just as in~\cref{sec:mxm1-queue:-expected}.

Consider the~$k$th job of some sample path of the $M/G/1$ queueing process.
Let the job's service start\sidenote{Usually this is later than the arrival time $A_{k}$.} at time $\tilde A_k$, so that it departs at time $D_k=\tilde A_k + S_k$, see~\cref{fig:mg1remainingservicetime}.
At time~$s$,  the remaining service time of job~$k$ is
$(D_k-s)\1{\tilde A_k \leq s < D_k}$.
\begin{marginfigure}
\begin{tikzpicture}[scale=1,
 open/.style={shape=circle, fill=white, inner sep=1pt, draw, node contents=},
 closed/.style={shape=circle, fill=black, inner sep=1pt, draw, node contents=}]
 \draw (-1,0) -- (4,0);
 %x\draw (1,0) -- (3,0) node[midway, fill=white] {$s$};
 \draw node (c1) at (0,3) [closed, label={}]
 node (c2) at (3,0)[open, label={}]
 (c1) to (c2);
 \draw[dotted] (0,0) -- (0,3) node[midway, rotate=90, fill=white] {$S_k$};
 \node[below] at (0,0) {$\tilde A_k$};
 \node[below] at (3,0) {$D_k$};

 \draw[<->, dotted] (0.75,0) -- (0.75,2.25) node[midway,fill=white,rotate=90] {$D_k-s$};
 \node[below] at (0.75,0) {$s$};
 \end{tikzpicture}

 \caption{Remaining service time.}
 \label{fig:mg1remainingservicetime}
\end{marginfigure}
By adding up all these times, we find that the total remaining service time as seen by the server up to time~$t$ is given by
 \begin{equation*}
 Y(t) = \int_0^t (D_{D(s)+1}-s)\1{A_{D(s)+1} \leq s \leq D_{D(s)+1}} \d s
 \end{equation*}
To see this, note that at time~$s$, the number of departures is $D(s)$.
 Thus, $D(s)+1$ is the first job to depart after time~$s$.
 The departure time of this job is $D_{D(s)+1}$, hence the remaining service time at time~$s$ is $D_{D(s)+1}-s$, provided this job is in service.
Then, as $t \to \infty$, $Y(t)/t$ converges to the  (time-average) remaining service time $\E{S_r}$.

 We also see in \cref{fig:mg1remainingservicetime} that $X_k = Y(D_k) - Y(D_{k-1})$ is the area under the triangle.
 By choosing $T_k=D_k$ as the epochs to inspect $Y(\cdot)$, $X=\lim_{n\to\infty} \frac{1}{n}\sum_{k=1}^n S_k^2/ 2 = \E{S^2}/2$.
 By the renewal-reward theorem, $Y=\delta X$, but $\delta = \lambda$,\sidenote{by rate-stability} and therefore
\begin{equation*}
\E{S_r} = \frac{\lambda}2 \E{S^2}.
\end{equation*}
Replacing this expression in the above expression for $\E\W$, we obtain the \recall{Pollaczek-Khinchine formula}
\begin{equation*}
 \E{\W} = \frac 1 2 \frac{\lambda \E{S^2}}{1-\rho} =\frac{1 + C_s^2}2 \frac{\rho}{1-\rho} \E S,
\end{equation*}
where the second equation follows from  the observation that
\begin{equation*}
\frac{\E{S^2}}{(\E S)^2} =  \frac{\E{S^2} - (\E S)^{2} + (\E{S})^{2}}{(\E S)^2} = \frac{\V S + (\E S)^{2}}{(\E S)^{2}} = C_s^2 +1.
\end{equation*}



\begin{exercise}\label{ex:l-241}
 Consider a workstation with just one machine.
 We model the job arrival process as a Poisson process with rate $\lambda=3$ per day.
 The average service time $\E S = 2$ hours, $C^2_s = 1/2$, and the shop is open for 8 hours per day.
Show that  $\E{\W} = 4.5h$. What would you propose to reduce $\E \W$ to 2h?
\begin{solution}
$\rho = \lambda \E S = (3/8)\cdot 2 = 3/4$, $\E{\W} = 4.5$ h.
If we were able to reduce all service variability, i.e., $C_s^2=0$, then still $\E \W = 3$h.
Hence, we have to increase capacity, or reduce $\E S$.
Another possibility is to plan the arrival of jobs such that $C_a^2=0$.
However, typically this is not possible.
Would you accept that the supermarket plans your visits?
\end{solution}
\end{exercise}





\begin{exercise}\label{ex:5}
Show for the $M/G/1$ that $\E{S_r} = \rho \E{S_r\given S_r>0}$.  As consequence, for the $M/M/1$ queue, $\E{S_{r}} \neq \E S$; why not?
\begin{hint}
Use the PASTA property. Realize that when estimating $\E{S_r}$ along a sample path, $S_r=0$ for jobs that arrive at an empty system.
\end{hint}
\begin{solution}
 The probability to find the server busy upon arrival is $\rho$, and only jobs that find the server occupied see a positive remaining service time $S_r>0$.
\begin{equation*}
\E{S_r} = \rho \E{S_r\given S_r >0} + (1-\rho) \E{S_r\given S_r = 0} = \rho \E{S_r\given S_r>0}.
\end{equation*}
For the $M/M/1$, service times are memoryless, hence, $\E{S_r \given S_r>0} = \E S$, and this implies that $\E{S_r} = \rho \E S$ for the $M/M/1$ queue.
\end{solution}
\end{exercise}



\begin{exercise}
Compare  the expected waiting time of the $M/D/1$ queue to that of the $M/M/1$ queue. %$\E{\W(M/D/1)}$ to $\E{\W(M/M/1)}$.
\begin{hint}
Use that $\V S=0$ for the $M/D/1$ queue and $C_{s}^{2} = 1$ for the $M/M/1$ queue.
\end{hint}
\begin{solution}
$\V S = 0 \implies C_s^2 = 0 \implies \E{\W_{M/D/1}} = {\E{\W_{M/M/1}}}/2$.
\end{solution}
\end{exercise}

\begin{exercise}
 Compute $\E\J$ for the $M/G/1$ queue with $S\sim U[0,\alpha]$.
\begin{solution}
 \begin{align*}
\E S &= \alpha/2, & \E{S^2} &= \int_0^\alpha x^2 \d x/\alpha = \alpha^2/3,\\
\V S &= \alpha^2/3 - \alpha^2/4= \alpha^2/12, & C_s^2 &= (\alpha^2/12)/(\alpha^2/4) = 1/3,\\
\rho &= \lambda \alpha/2,\\
\E{\W} &= \frac{1+C_s^2}2 \frac{\lambda \alpha/2}{1-\lambda \alpha/2}\frac \alpha2, &
\E \J &= \E{\W} + \frac \alpha2.
 \end{align*}
\end{solution}
\end{exercise}


\begin{exercise}
Find an expression in terms of $\lambda$ and $\E S$ for the acceptance and loss probabilities of  the $M/G/1/1$ queue.
Why is this expression not valid for the $G/G/1/1$ queue.
\begin{hint}
 The rate of accepted jobs is $\lambda \pi(0)$, hence $\rho= \lambda \pi(0) \E S$.
But also $\rho = 1-\pi(0)$.  Now solve for $\pi(0)$.
\end{hint}
\begin{solution}
 \begin{equation*}
 \lambda\pi(0) \E S = 1 - \pi(0) \iff \pi(0)=\frac1{1+\lambda\E S}
\iff 1-\pi(0) = \frac{\lambda \E S}{1+\lambda \E S}.
 \end{equation*}

 There is another way to derive this.
 The system can contain at most 1 job.
 Necessarily, if the system contains a job, this job must be in service.
 All jobs that arrive while the server is busy are rejected.
 Just after a departure, the average time until the next arrival is $1/\lambda$, and then a new service starts with an average duration of $\E S$.
 After this departure, a new cycle starts.
 Thus, $\rho = \E S/(1/\lambda + \E S) = \lambda \E S/ (1 + \lambda \E S)$.


For the $G/G/1/1$ queue the PASTA property does not hold, hence $p(0)\neq \pi(0)$ in general.
\end{solution}
\end{exercise}


\input{trailer}
