\documentclass[queueing-book.tex]{subfiles}
\input{header}

\begin{document}

\section{Simulating Inventory Systems}
\label{sec:simul-invent-syst}

In this section we first develop a simulation environment to analyze an inventory system.

\newthought{The Lighthouse Company} sells ceiling lamps and considers to use an $(s,S)$ policy to control the inventory since the replenishment cost, $K=50$ euro is relatively high. The company wants to use simulation to find sound values for~$s$  and~$S$.

The details of the case are like this.
Ceiling lamps sell at 100 Euro while the buying price is 40 Euro.
The monthly holding cost is estimated to be $50\%$ of the buying price.
To prevent customers from buying a lamp at a competitor (e.g., Internet sales), the Lighthouse Company offers a reduction of $20\%$ per day of the regular selling price when customers cannot be satisfied from on-hand stock.
The pmf~$f$ of the daily demand is given by
  \begin{align*}
    f_{0} &= 1/6, &   f_{1} &= 1/5, &  f_{2}&= 1/4, &
    f_{3} &= 1/8, &  f_{4} &= 11/120, &  f_{5} &= 1/6.
  \end{align*}
The leadtime $L=2$ days.

With the code below we set up a simulation to estimate the performance measures for one specific $(s,S)$ policy.
We go step by step, and provide explanations above the code to which it relates.\sidenote{You should use Google, or ChatGPT, to look up further information if necessary.}. You can copy the code to a file and it should run without problems (provided you copy all lines correctly).



Most of the next modules should by now be familiar to you.
\pyth/scipy.stats/ offers a nice class \pyth/rv_discete/ that we can use, among much other functionality, to generate random deviates of specific distribution.
Finally, \pyth/icecream/ provides some more overview of what is printed than the regular \pyth/print()/ function.
\inputminted[firstline=2, lastline=7]{python}{../code/ss_inventory_simulation.py} % modules

The documentation of \pyth/rv_discrete/ that to make a distribution we have to provide two arrays of numbers, the first is the support of the rv, the second the pmf.
\inputminted[firstline=11, lastline=13]{python}{../code/ss_inventory_simulation.py} % rvs



\noindent Here are the other model parameters.
Check in particular how we compute the cost parameters.\sidenote{Getting the units right is not always easy.}

You should also check the line numbers at the left margins in the code blocks.
Where the line numbers skip from one block to the next, the code in my python file has empty lines.

\inputminted[firstline=17, lastline=22]{python}{../code/ss_inventory_simulation.py} % data

\noindent
We next compute the inventory positions and inventory levels.
For ease we combine ~\cref{eq:i5} into single recursion to update $P_t'$, and likewise for $I_t'$.
In the code, we write \pyth/Pp/ to say `Pprime', to recall that we mean the end-of-period inventory position.
With regard to line 27, you should know that expressions like \pyth/(3 <= 4)/ evalutate to \pyth/True/ which in turn become~$1$ in computations. Thus, \pyth/Pp[t - 1] <= s)/ implements the indicator functions.\sidenote{Ask ChatGPT to explain such code, it's a great help.}
Thus, We generate \pyth/N/ random deviates for the demand and use the random generator to ensure we get the same numbers every time we run the simulation.
The first \pyth/for/ loop for the inventory levels computes $\IL$ for the first couple of period, because $t-L< 0$ for $t< L$.
In the second loop we should protect against a simple error: it might happen that the number of periods over which we simule is smaller than~$L$.
\inputminted[firstline=26, lastline=41]{python}{../code/ss_inventory_simulation.py} % simulation

\noindent
The last block of code computes the measures.
The cycle service level~$\alpha_{c}$ is worth to study.\sidenote{Ask ChatGPT for explanation if you find this line hard.}
We use slicing, which is a technique that is much used in python, but also in matlab.
\inputminted[firstline=45, lastline=58]{python}{../code/ss_inventory_simulation.py} % results


It remains to find good policy parameters, i.e., $s$ and~$S$.
In principle this is simple: just change~$s$ and~$S$ over a range of values and see how that affects the performance measures.
However, finding the overal \emph{optimal} parameters \emph{efficiently} is not easy.
To minimize the cost, I have published an article on how to do that, but I did not study how to minimize average cost under some constraint on one of the service level measures, for instance.



\input{trailer}
